---
# title: "The document title"
output:
  bookdown::pdf_document2:
    toc: false
    includes:
      in_header: "preamble.tex"
    latex_engine: lualatex
  word_document:
    reference_docx: style_reference.docx
always_allow_html: true
bibliography: bibliography.bib
csl: nlm.csl
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, results = "hide", message = FALSE, warning = FALSE
)
```

```{r source}
library(here)
library(broman)
library(kableExtra)
source(here("code", "02_explore.R"))
```

<!--
\begin{center}
Lucas Veras$^1$, Florêncio Diniz-Sousa$^1$, Giorjines Boppre$^1$, Vítor Devezas$^2$, Hugo Santos-Sousa$^2$, John Preto$^2$, João Paulo Vilas-Boas$^{3,4}$, Leandro Machado$^{3,4}$, José Oliveira$^1$, Hélder Fonseca$^1$
\end{center}

\bigskip

1 - Research Center in Physical Activity, Health and Leisure (CIAFEL), Faculty of Sport, University of Porto, Porto, Portugal

2 - General Surgery Department, São João Medical Center, Porto, Portugal

3 - Center of Research, Education, Innovation and Intervention in Sport (CIFI2D), Faculty of Sport, University of Porto, Porto, Portugal

4 - Biomechanics Laboratory (LABIOMEP-UP), University of Porto, Porto, Portugal
-->

# Introduction {.unnumbered}

The aim of this study was to develop pGRF and pLR prediction models based on accelerometer data during walking and running.

# Methods {.unnumbered}

## Participants {.unnumbered}

A convenience sample of `r max(sample_size$pGRF_N)` adults (`r unname(sample_size_sex["M"])` males; $`r sample_descriptives$age_mean`\pm`r sample_descriptives$age_sd`$ years; $`r sample_descriptives$height_mean`\pm`r sample_descriptives$height_sd`$ cm; $`r sample_descriptives$body_mass_mean`\pm`r sample_descriptives$body_mass_sd`$ kg; $`r sample_descriptives$BMI_mean`\pm`r sample_descriptives$BMI_sd`$ $\mathrm{kg\cdot m^{-2}}$, $\overline{X}\pm SD$) with no self-reported neurological or musculoskeletal limitations was recruited. Height and body mass were assessed following standard procedures with a stadiometer and a digital scale, respectively. All participants were informed about the experiments purpose and protocol before giving written informed consent. The study was approved by the local Ethics Committee (CES 192-14).

## Protocol {.unnumbered}

Data collection took place at Porto Biomechanics Laboratory (LABIOMEP-UP) in an instrumented treadmill with built-in FP (AMTI Corporation, Watertown, USA). Before starting the data collection, participants were given time to warm up on the treadmill. The protocol consisted of walking and running in 13 incremental speeds, with 0º inclination, from 2 to 14 $\mathrm{km\cdot h^{-1}}$ with 1 $\mathrm{km\cdot h^{-1}}$ increment. Walking and running speeds ranged from 2 to 6 $\mathrm{km\cdot h^{-1}}$ and from 7 to 14 $\mathrm{km\cdot h^{-1}}$, respectively. Each walking speed lasted for 1 minute and each running speed for 30 seconds. To ensure single foot contact on each FP during the trial, feedback regarding location on the treadmill was provided by a researcher. Participants wore their own sports shoes throughout the trials.

\medskip

During the protocol, participants wore three activity monitors, one at their right hip (along the anterior axillary line, at the level of the iliac crest), another at their lower back (at the midpoint between the two posterior superior iliac spines) and the last at their right ankle (immediately superior to the lateral malleolus). Activity monitors at waist level were secured on the same elastic belt with a clip and the activity monitor placed on the ankle was secured by an elastic belt fixed with adhesive tape. In all placements the accelerometer y-axis (vertical) was aligned with the standing body longitudinal axis. All participants wore the same three activity monitors always positioned at the same places.

\medskip

Activity monitors used were the GT9X Link model ($\mathrm{\pm16}$\textit{g} range; ActiGraph, Pensacola, USA) that incorporates primary and secondary triaxial accelerometers. As a manufacturer proprietary filter is applied on the primary, but not on the secondary accelerometer raw data, only data from the latter was used in this study, reducing filtering bias and enhancing the data processing replicability.

\medskip

Data from FP and accelerometers were collected at 1000 Hz and 100 Hz sampling frequency, respectively, and operated through the manufacturer supplied software (for the FP: Netforce, Version 3.5.1; AMTI Corporation, Watertown, USA; for the accelerometers: ActiLife version 6.13.3; Actigraph, Pensacola, USA). Raw GRF (expressed in N) and ACC (expressed in gravitational ACC units; 1\textit{g} = 9.807 $\mathrm{m\cdot s^{-2}}$) from the x, y and z vectors were exported to plain text files.

## Data processing {.unnumbered}

FP and accelerometer data were processed using MATLAB (Version 8.3, Mathworks, Natick, USA), as described elsewhere [@Veras2020]. Briefly, both GRF and ACC signals were filtered using a Butterworth fourth-order low-pass filter, with 20 Hz cut-off frequency, and their resultant vector calculated ($r_i = \sqrt{x_i^2 + y_i^2 + z_i^2}$). After that, peak ACC (pACC) values were defined as minimum height of twice the positive ACC average, during any given speed, and separated by 0.4 seconds minimum. GRF and ACC signals were adjusted through the time set by the systems clock, with manual correction based on visual inspection for possible errors, and then synchronized by the maximum cross-correlation coefficient. Peak GRF (pGRF) was determined as the highest value between 0.4 seconds before and after each pACC. The rates of change, LR for the GRF and ACC rate (AR) for the ACC, were computed through a second-order centered derivative from the beginning of the foot contact to the pGRF, and pLR and peak AR (pAR) were defined as the maximum value in the array.

\[f_i^{'} = \frac{f_{i + 1} - f_{i - 1}}{t_{i + 1} - t_{i - 1}}\]

\[f_{i_{max}}^{'} = max(f_i^{'})\]

Regions of interest for data analysis were manually selected where locomotion patterns were constant, which corresponded of 20-45 seconds periods, on average. Then, pGRF, pACC, pLR and pAR means of the resultant and its vertical component for each participant at each speed were extracted and used in all remaining analyses. In walking trials, a median of `r n_peaks_desc[which(n_peaks_desc$locomotion_type == "walking"), "n_peaks_median"]` peaks (interquartile range `r n_peaks_desc[which(n_peaks_desc$locomotion_type == "walking"), "n_peaks_median"] - n_peaks_desc[which(n_peaks_desc$locomotion_type == "walking"), "n_peaks_iqr"]`\textendash`r n_peaks_desc[which(n_peaks_desc$locomotion_type == "walking"), "n_peaks_median"] + n_peaks_desc[which(n_peaks_desc$locomotion_type == "walking"), "n_peaks_iqr"]`) were used to calculate these means, while in running trials it was a median of `r n_peaks_desc[which(n_peaks_desc$locomotion_type == "running"), "n_peaks_median"]` peaks (interquartile range `r n_peaks_desc[which(n_peaks_desc$locomotion_type == "running"), "n_peaks_median"] - n_peaks_desc[which(n_peaks_desc$locomotion_type == "running"), "n_peaks_iqr"]`\textendash`r n_peaks_desc[which(n_peaks_desc$locomotion_type == "running"), "n_peaks_median"] + n_peaks_desc[which(n_peaks_desc$locomotion_type == "running"), "n_peaks_iqr"]`).

## Statistical analyses {.unnumbered}

Statistical analyses were conducted using R statistical software (version `r paste(R.Version()$major, R.Version()$minor, sep = ".")`, R Foundation for Statistical Computing, Vienna, Austria). All statistical analyses were registered in an open platform, where R code utilized and detailed information can be assessed (LINK). The statistically significant value was set as $\alpha$ = 0.05.

\medskip

Regression equations to predict pRGFR, pVGRF, pRLR and pVLR were developed by linear mixed models. Body mass (kg) was tested as a predictor in all models. In addition, in the pRGRF and pVGRF models, pRACC and peak vertical ACC (pVACC; \textit{g}), respectively, were tested as predictors, while in the pRLR and pVLR models, peak resultant ATR (pRATR; $g\phantom{ }\mathrm{\phantom{ }\cdot s^{-1}}$) and peak vertical ATR (pVATR; $g\phantom{ }\mathrm{\phantom{ }\cdot s^{-1}}$), respectively, were tested as predictors. All predictors were tested as fixed effects and were shown to be significant. Subject and speed were tested as random effects, and the inclusion of both has showed to improve the models. Final models were chosen according to -2 log-likelihood statistics [@Field2012]. Traditional coefficient of determination ($\mathrm{R^2}$) was represented by conditional $\mathrm{R^2}$, that estimates the variance explained by the whole model [@Nakagawa2013].

\medskip

Model was validated using the leave-one-out cross-validation approach [@Staudenmayer2012]. For leave-one-out cross-validation each participant’s data was separated into a testing dataset (one participant at a time) with the remaining data in the training dataset. New linear mixed models with the same outcomes and predictors as determined for the entire sample were developed using the training dataset and then used to predict the outcome for the participant in the testing dataset. This process was repeated for all participants (`r max(sample_size$pGRF_N)` times). Data from the testing dataset was used in the remaining statistical analysis.

\medskip

To assess the models prediction accuracy, mean absolute error (MAE), mean absolute percent error (MAPE) and root mean square error (RMSE) were calculated. Also, methods agreement was evaluated with Bland-Altman plots by plotting the difference between the actual (FP measured) and predicted values against their mean. Bias was computed as the mean of these differences and the limits of agreement were obtained using \pm 1.96 standard deviation of the mean between actual and predicted values [@Bland1986]. In order to test whether there was a consistent over or underestimation of the acual values, one-sample \textit{t} tests were performed to assess if the difference between actual and predicted values are different than zero. Linear regressions were applied to identify if the magnitude of the mean between actual and predicted values influenced the magnitude of their difference. This was done to ascertain if the prediction accuracy was constant throughout the assayed magnitude range [@Giavarina2015]. These analyses were conducted separately for data from each accelerometer placement and outcome.

\medskip

Finally, our pGRF prediction equation was compared with a previously published reference equation [@Neugebauer2014], that was developed using similar methods. This compaarison was performed in three ways using: i) the whole sample; ii) a subsample of walking trials; iii) a subsample of running trials. To assess the prediction accuracy, MAE, MAPE and RMSE were calculated using pVGRF values predicted from both equations. No pLR prediction equation suited to be compared with ours was found.

\medskip

# Results {.unnumbered}

```{r bmi_class}
BMI_avg <- sample_descriptives$BMI_mean
BMI_class <- NA
if (BMI_avg >= 18.5 & BMI_avg < 25) {
  BMI_class <- "normal weight"
} else if (BMI_avg >= 25 & BMI_avg < 30) {
  BMI_class <- "overweight"
} else if (BMI_avg >= 30 & BMI_avg < 35) {
  BMI_class <- "obesity class I"
} else if (BMI_avg >= 35 & BMI_avg < 40) {
  BMI_class <- "obesity class II"
} else if (BMI_avg >= 40) {
  BMI_class <- "obesity class III"
}
```
Our recruited sample presented an average BMI of `r sample_descriptives$BMI_mean` $\mathrm{kg\cdot m^{-2}}$, which is relatively high and classified as `r BMI_class`. 
It would be expected that participants with higher BMI values, due to low motor ability and physical capacity, are able to perform the trials only at the lowest speeds.
This is precisely what happened and, as can be observed in Figure 1, from 2 to 5 $\mathrm{km\cdot h^{-1}}$ the BMI median is around 30 $\mathrm{kg\cdot m^{-2}}$, decreasing to around 27 $\mathrm{kg\cdot m^{-2}}$ at 6 $\mathrm{km\cdot h^{-1}}$, which is the fastest walking speed, with a further reduction to less than 25 $\mathrm{kg\cdot m^{-2}}$ at the running speeds. Furthermore, only one participant with a BMI higher than 30 $\mathrm{kg\cdot m^{-2}}$ was able to run. This implies in a sample size reduction from `r filter(sample_size, acc_placement == "hip" & vector == "resultant" & speed == "2")$pGRF_N` at the slowest speed to `r filter(sample_size, acc_placement == "hip" & vector == "resultant" & speed == "14")$pGRF_N` at the fastest.

\begin{figure}[ht]
	\captionsetup{singlelinecheck = false, format = hang, justification = raggedright, font = footnotesize, labelsep = space}
	\centering
	\begin{measuredfigure}
		\includegraphics[width=\textwidth]{fig1.png}
		\caption{Distribution of participant’s body mass index per speed. Dots represent individual participants.}
	\end{measuredfigure}
\end{figure}

\medskip

As the main variables collected in this experiment are the peak magnitudes and rates of the GRF and ACC signals, there was a need to first establish their relationship. During locomotion, pACC increases linearly as pGRF increases. Such relation can be observed in the Figure 2, Panel A, for from hip-worn accelerometers and the resultant vector, and in Supplemental Figure S1 for all placements and the resultant vector and its vertical component. Moreover, for the same pACC values, the registered pGRF tended to be higher in subjects with higher BMI. It demonstrates that accelerometer data can adequately discriminate between different BMI categories. However, data from the pAR x pLR relationship show a greater dispersion, especially at higher values, and cannot discriminate so well the different BMI categories, as can be observed in Figure 2, Panel B, and Supplemental Figure S2.

\begin{figure}[ht]
	\captionsetup{singlelinecheck = false, format = hang, justification = raggedright, font = footnotesize, labelsep = space}
	\centering
	\begin{measuredfigure}
		\includegraphics[width=\textwidth]{fig2.png}
		\caption{Relationship between variables collected by force plates and accelerometers worn at hip, for the resultant vector. Linear trends for each body mass index category are also depicted. Panel A shows peak magnitudes (pRGRF and pRACC) while Panel B shows peak rates (pRLR and pRAR). Abbreviations: pRACC, peak resultant acceleration; pRAR, peak resultant acceleration rate; pRGRF, peak resultant ground reaction force; pRLR, peak resultant loading rate.}
	\end{measuredfigure}
\end{figure}

\medskip

```{r GRF_accuracy}
load(here("output", "prediction_models.rda"))
res_GRF_accuracy <- purrr::map(
  accuracy_res_GRF_models,
  ~ dplyr::mutate(.x, MAPE = as.numeric(MAPE))
) %>%
  purrr::map_dfr(rbind) %>%
  dplyr::mutate(
    placement = c("ankle", "lower back", "hip"),
    vector = "resultant"
  )
ver_GRF_accuracy <- purrr::map(
  accuracy_ver_GRF_models,
  ~ dplyr::mutate(.x, MAPE = as.numeric(MAPE))
) %>%
  purrr::map_dfr(rbind) %>%
  dplyr::mutate(
    placement = c("ankle", "lower back", "hip"),
    vector = "vertical"
  )
GRF_accuracy <- rbind(res_GRF_accuracy, ver_GRF_accuracy)
hip_GRF_accuracy <- GRF_accuracy %>%
  filter(placement == "hip" & vector == "resultant")
```

Regression equations were developed to predict the mechanical loading variables based on accelerometry data. In the pGRF prediction equations, body mass and pACC were significant predictors and included in the models. Table 1 shows the regression coefficients, $\mathrm{R^2}$ and accuracy indices from models derived from accelerometers placed at ankle, lower back and hip, and from the resultant vector and its vertical component. In all pGRF prediction models, $\mathrm{R^2}$ ranged from `r min(GRF_accuracy$R2_cond)` to `r max(GRF_accuracy$R2_cond)`, which shows that the predictor variables are able to explain in a great extent the outcome. Also, MAE ranged between `r myround(min(GRF_accuracy$MAE), 1)` N and `r myround(max(GRF_accuracy$MAE), 1)` N, and the MAPE from `r myround(min(GRF_accuracy$MAPE) * 100, 1)`\% to `r myround(max(GRF_accuracy$MAPE) * 100, 1)`\%, which is a smaller error than previously published equations [@Neugebauer2014; @Fortune2014], and similar to the error from equations previously published by out group, but which included walking only [@Veras2020]. RMSE were also computed and shown to range between `r myround(min(GRF_accuracy$RMSE), 1)` N and `r myround(max(GRF_accuracy$RMSE), 1)` N. Of all the pGRF predictions equations developed, the one based on hip-worn accelerometers and the resultant vector has shown the best performance, with an $\mathrm{R^2}$ of `r hip_GRF_accuracy$R2_cond`, and the lowest MAE (`r myround(hip_GRF_accuracy$MAE, 1)` N), MAPE (`r myround(hip_GRF_accuracy$MAPE * 100, 1)`\%) and RMSE (`r myround(hip_GRF_accuracy$RMSE, 1)` N).

\medskip

```{r LR_accuracy}
load(here("output", "prediction_models.rda"))
res_LR_accuracy <- purrr::map(
  accuracy_res_LR_models,
  ~ dplyr::mutate(.x, MAPE = as.numeric(MAPE))
) %>%
  purrr::map_dfr(rbind) %>%
  dplyr::mutate(
    placement = c("lower back", "hip"),
    vector = "resultant"
  )
ver_LR_accuracy <- purrr::map(
  accuracy_ver_LR_models,
  ~ dplyr::mutate(.x, MAPE = as.numeric(MAPE))
) %>%
  purrr::map_dfr(rbind) %>%
  dplyr::mutate(
    placement = c("lower back", "hip"),
    vector = "vertical"
  )
LR_accuracy <- rbind(res_LR_accuracy, ver_LR_accuracy)
back_LR_accuracy <- LR_accuracy %>%
  filter(placement == "lower back" & vector == "resultant")
```

Moreover, pLR prediction equations were developed, and these included as predictors the pAR and again body mass. Regression coefficients, $\mathrm{R^2}$ and accuracy indices for these equations are also shown in Table 1. For these equations, the lowest $\mathrm{R^2}$ found was `r min(LR_accuracy$R2_cond)` and the highest was `r max(LR_accuracy$R2_cond)`, reflecting a moderate prediction ability. MAE, MAPE and RMSE values ranged from `r myround(min(LR_accuracy$MAE), 1)` N to `r myround(max(LR_accuracy$MAE), 1)` N, from `r myround(min(LR_accuracy$MAPE) * 100, 1)`\% to `r myround(max(LR_accuracy$MAPE) * 100, 1)`\%, and from `r myround(min(LR_accuracy$RMSE), 1)` N to `r myround(max(LR_accuracy$RMSE), 1)` N, respectively. The results found for the pLR prediction equations were all considerably poorer than those for the pGRF equations. The accuracy indices values were also somewhat worse than previously published pLR prediction equations [@Veras2020; @Fortune2014], although equations from both studies were only validate for walking. The best-performing pLR prediction equation was the one for accelerometers worn at lower back and the resultant vector, with an $\mathrm{R^2}$ of `r back_LR_accuracy$R2_cond` and a MAE, MAPE and RMSE of `r myround(back_LR_accuracy$MAE, 1)` N, `r myround(back_LR_accuracy$MAPE * 100, 1)`\% and `r myround(back_LR_accuracy$RMSE, 1)` N, respectively.

\medskip

```{r accuracy_table, results = "show"}
source(here("tables", "tab1.R"))
tab1
```

```{r ba_regression}
source(here("code", "04_sub_analyses.R"))
```

Afterwards, Bland-Altman plots were built to visualize the agreement between the actual and predicted pGRF and pLR, for both the resultant vector and its vertical component and all accelerometer placements. A good agreement in a Bland-Altman plot is observed when values tend to cluster around zero, showing no consistent over or underestimation of the actual values. Also, the majority of data points should lie within the limits of agreement, which are determined by \pm 1.96 standard deviation of the bias (average difference between actual and predicted values). Bland-Altman plots for the pRGRF and pRLR from hip-worn accelerometers can be seen in Figure 3, and for all accelerometers placements and both resultant and vertical vectors in the Supplemental Figures S3 and S4. Values for all pGRF and pLR prediction equations developed tended to assemble around zero and between the limits of agreement. For the pGRF equations, a tendency to higher dispersion can be observed in the higher magnitudes, which correspond to the running trials, and especially in values derived from ankle-worn accelerometers (Supplemental Figure S3, Panels A and D). Prediction equations for pLR showed a more accentuated dispersion, and again in the higher magnitudes. None of the regression models presented a bias different than zero ($p > 0.05$), demonstrating that these models do not consistently over or underestimate the pGRF and pLR values. Furthermore, linear regressions were applied to identify proportional bias, that is, to test whether pGRF and pLR prediction errors were systematically influenced by their magnitude. The pGRF prediction equation based on ankle-worn accelerometers data for the resultant vector showed a significant proportional bias ($p = `r myround(coefficients(ba_regression_res_GRF$ankle)[2, 4], 2)`$), although with a low magnitude ($\mathrm{R^2} = `r myround(ba_regression_res_GRF$ankle["adj.r.squared"], 2)`$).

\begin{figure}[ht]
	\captionsetup{singlelinecheck = false, format = hang, justification = raggedright, font = footnotesize, labelsep = space}
	\centering
	\begin{measuredfigure}
		\includegraphics[width=\textwidth]{fig3.png}
		\caption{Bland-Altman plots showing agreement between actual and predicted pRGRF (Panel A) and pRLR (Panel B) for hip-worn accelerometers. Continuous thick lines show bias (average of the differences between actual and predicted values) while dashed lines show the limits of agreement (bias \pm 1.96 standard deviation). Abbreviations: pRGRF, peak resultant ground reaction force; pRLR, peak resultant loading rate.}
	\end{measuredfigure}
\end{figure}

\medskip

Lastly, we compared the performance of our equations with a reference equation previously published. As the only comparable equation found was the one to predict pVGRF using hip-worn accelerometer data developed by Neugebauer et al [@Neugebauer2014], we applied it to our data and computed the MAE, MAPE and RMSE. Then, these indices were compared to the indices for our equation (Table 1). Apart from the overall equations performance, it was done a sub-analysis separating the walking and running trials. The results of these comparisons can be observed in the Table 2. Accuracy indices of our equations were all remarkably better compared to the reference equation, with a MAPE approximately 2 to 3 times lower. Furthermore, our equation performance was constant between walking and running, indicating that it can be used for both activities without loss of accuracy.

```{r comparison_table, results = "show"}
source(here("tables", "tab2.R"))
tab2
```

\newpage

# References {.unnumbered}

<div id="refs"></div>

\newpage

# Supplemental Material {.unnumbered}

## Supplemental Figure S1 {.unnumbered}
\begin{figure}[ht]
	\captionsetup{singlelinecheck = false, format = hang, justification = raggedright, font = footnotesize, labelsep = space}
	\centering
	\begin{measuredfigure}
		\includegraphics[width=\textwidth]{figS1.png}
	\end{measuredfigure}
\end{figure}
\footnotesize Relationship between peak ground reaction force and acceleration magnitudes. Data from accelerometers worn at ankle (Panels A and D), lower back (Panels B and E) and hip (Panels C and F), for both the resultant vector (Panels A to C) and its vertical component (Panels D to F). Linear trends for each body mass index category are also depicted. Abbreviations: pRACC, peak resultant acceleration; pRGRF, peak resultant ground reaction force; pVACC, peak vertical acceleration; pVGRF, peak vertical ground reaction force.

\medskip

## Supplemental Figure S2 {.unnumbered}
\begin{figure}[ht]
	\captionsetup{singlelinecheck = false, format = hang, justification = raggedright, font = footnotesize, labelsep = space}
	\centering
	\begin{measuredfigure}
		\includegraphics[width=\textwidth]{figS2.png}
	\end{measuredfigure}
\end{figure}
\footnotesize Relationship between peak loading and acceleration rates. Data from accelerometers worn at ankle (Panels A and D), lower back (Panels B and E) and hip (Panels C and F), for both the resultant vector (Panels A to C) and its vertical component (Panels D to F). Linear trends for each body mass index category are also depicted. Abbreviations: pRAR, peak resultant acceleration rate; pRLR, peak resultant loading rate; pVAR, peak vertical acceleration rate; pVLR, peak vertical loading rate.

\newpage

## Supplemental Figure S3 {.unnumbered}
\begin{figure}[ht]
	\captionsetup{singlelinecheck = false, format = hang, justification = raggedright, font = footnotesize, labelsep = space}
	\centering
	\begin{measuredfigure}
		\includegraphics[width=\textwidth]{figS3.png}
	\end{measuredfigure}
\end{figure}
\footnotesize Bland-Altman plots showing agreement between actual and predicted pGRF for accelerometers worn at ankle (Panels A and D), lower back (Panels B and E) and hip (Panels C and F), for both the resultant vector (Panels A to C) and its vertical component (Panels D to F). Continuous thick lines show bias (average of the differences between actual and predicted values) while dashed lines show the limits of agreement (bias \pm 1.96 standard deviation). Abbreviations: pGRF, peak ground reaction force; pRGRF, peak resultant ground reaction force; pVGRF, peak vertical ground reaction force.

## Supplemental Figure S4 {.unnumbered}
\begin{figure}[ht]
	\captionsetup{singlelinecheck = false, format = hang, justification = raggedright, font = footnotesize, labelsep = space}
	\centering
	\begin{measuredfigure}
		\includegraphics[width=\textwidth]{figS4.png}
	\end{measuredfigure}
\end{figure}
\footnotesize Bland-Altman plots showing agreement between actual and predicted pLR for accelerometers worn at ankle (Panels A and D), lower back (Panels B and E) and hip (Panels C and F), for both the resultant vector (Panels A to C) and its vertical component (Panels D to F). Continuous thick lines show bias (average of the differences between actual and predicted values) while dashed lines show the limits of agreement (bias \pm 1.96 standard deviation). Abbreviations: pLR, peak loading rate; pRLR, peak resultant loading rate; pVLR, peak vertical loading rate.
